Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-01-16 LaTeX/Turnover.tex
changed to cis-GSD and trans-GSD
Sally edits to intro and discussion. I also removed the discussion of two-locus drive and added something about ESD to GSD transitions (which we don't cover) and incomplete masculinizers/feminizers

%involve sex-ratio selection or sex-specific selection (e.g., sexually antagonistic selection).

@@

(1) We explicitly consider how selection on very tightly sex-linked loci influences the spread of novel sex-determiners. 
We find that tightly sex-linked genetic variation can favour the spread of new sex-determination systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or unlinked) to the sex-linked locus under selection, which is not predicted by previous theory. 

@@

We find that associations with haploid selected loci can drive transitions between sex determination systems, as has been shown for sexually-antagonistic loci. 
In addition, with haploid selection, transitions between male and female heterogamety can also evolve when linkage with the sex-determining locus is weakened. 
Haploid selection in the heterogametic sex can cause sex ratio biases, which may increase or decrease with the spread of new sex chromosomes. 
Thus, transitions between sex-determination systems cannot be simply predicted by selection to equalise the sex-ratio. 

@@

%A surprising result given that other determinants of sex allocation typically experience strong Fisherian sex-ratio selection to equalize sex ratios. 

@@

Overall, our models reveal that transitions between sex-determination systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unexpected genomic locations that experience selection during diploid and/or haploid phases.
These results might be reflected in the lability with which sex-determination systems evolve. %and for the factors determining the phylogenetic pattern of sex-determination systems. 

@@

Predominant theories accounting for the spread of new sex-determination systems by selection involve fitness differences between sexes (e.g., sexually antagonistic selection) or sex-ratio selection \citep{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
\citet{vanDoorn:2007eu,vanDoorn:2010hu} show that new sex-determining loci can be favoured if they arise in closer linkage with a locus that experiences sexual antagonism. 
Tighter linkage allows a stronger favourable association to build up between a male-beneficial allele, and a neo-Y chromosome, for example. 
Such associations can favour a new master sex-determining gene on a new chromosome \citep{vanDoorn:2007eu} and can also favour a transition between male and female heterogamety \citep[e.g., a ZW to XY transition,][]{vanDoorn:2010hu}. 
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determination locus will develop similar, favourable associations and hinder the spread of a new sex-determination system. 
%Here we extend these studies by explicitly calculating the the equilibrium allele frequencies of loci that are very tightly linked to the ancestral sex-determining region.

The sex ratio is directly determined by the sex-determination system, and it has therefore been suggested that sex-ratio selection is a dominant force in the evolution of sex determination (e.g., \citealt{Bull:1983vi}, p 66-67; \citealt{Beukeboom:2014vb}, Chapter 7). 
`Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce \citep{Fisher:1930wy,Charnov:1982wg}.
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female \citep{West:2009we}. 
Thus, if the population sex ratio is biased towards one sex, the average per-individual contribution of genetic material to the next generation from the opposite sex is greater. 
Therefore, a mutant that increases investment in the rarer sex will spread via the higher per-individual contributions made by that sex. 
%The default mode of sex-ratio selection is `Fisherian' sex-ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
%Given that the sex-determination system can directly affect the sex ratio, we might expect Fisherian sex-ratio selection to influence the spread of new sex-determination systems. 
In the case of sex-chromosome evolution, \citet{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased \citep{Hamilton:1967ts}. 
They find that new, unlinked sex-determining loci (masculinizing or feminizing mutations, i.e., neo-Y or neo-W loci) can then spread, which restore an even sex ratio.

@@

% from Bai et al. (2016) ``[Segregation distortion] phenomenon is commonly detected via genetic mapping and has been documented in various species, including mouse (Eversley et al., 2010; Casellas et al., 2012), Drosophila (Phadnis and Orr, 2009; Larracuente and Presgraves, 2012; McDermott and Noor, 2012), Tigriopus (Pritchard et al., 2011), rice (Koide et al., 2012; Reflinur et al., 2014; Xu et al., 2014), maize (Tang et al., 2013), and cotton (Yu et al., 2011; Hulse-Kemp et al., 2015).'' Also more in that paper that might be useful. 
%Drosophila http://www.biorxiv.org/content/early/2017/02/06/106336
Segregation distortion provides putative evidence of haploid selection and can sometimes be attributed to meiotic drive and/or gametic competition \citep{Lalanne2004,Fishman2005,Leppala2008,Leppala2013,Didion2015,Didion2016}.
Where it has been characterized, meiotic drive generally occurs either during the production of male or female gametes only \citep{Ubeda:2005gw,Lindholm:2016cw}.
Gametic competition is also typically sex specific, occurring primarily among male gametes, because there are typically many more pollen/sperm than required for fertilization.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes are known to cause a response to selection \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
Nevertheless, recent studies have demonstrated that sperm competition in animals can alter haploid allele frequencies and increase offspring fitness \citep{Immler:2014im,Alavioon2017}.

There are various ways by which genes experiencing haploid selection could influence transitions between sex-determination systems. 
If we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
In this respect, we might expect that haploid selection would affect transitions between sex-determination systems in a similar manner to sex-specific diploid selection \citep[as explored by][]{vanDoorn:2007eu,vanDoorn:2010hu}. 
That is, new masculinizing mutations (neo-Y chromosomes) could be favoured via associations with alleles that are beneficial in the male haploid stage. 
On the other hand, sex ratios can also become biased by linkage between the sex-determining region and a locus that harbours genetic variation in haploid fitness. 
For example, there are several known cases of sex-ratio bias caused by sex-linked meiotic drive alleles \citep[][Chapter 3]{Burt:2006} or selection among X- and Y-bearing pollen \citep{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
It is not immediately clear how the spread of new sex-determination systems would be influenced by the combination of sex-ratio biases and associations between haploid selected loci and sex-determining regions.

@@

We find that the spread of novel sex-determiners is influenced by both Fisherian sex-ratio selection and by selection on genetically-associated alleles. 
%We find that sex-ratio biases caused by haploid selection can exert Fisherian sex-ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex-ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
Surprisingly, Fisherian sex ratio selection does not dominate; it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing increases in sex-ratio bias. 
In addition to considering haploid selection, another novel development in our model is that we consider loci that are in very tight linkage with the ancestral sex-determining region. 
Because sex-determining loci are often found within a region of suppressed recombination, there can be a significant number of tightly linked loci. 
We find that loci linked with the ancestral sex-determining region can drive transitions in which the heterogametic sex changes, even when the neo-sex-determining locus is less closely linked to loci under selection (either including haploid selection or not). 
%Such transitions are not favoured in models lacking tight linkage and/or haploid selection. 
%Secondly, we allow sex-specific haploid selection to occur on a locus in tight or loose linkage with the ancestral sex-determining region. 
%In addition, when linkage between the ancestral sex-determining locus and a locus under haploid selection is tight, sex ratio biases can be strong such that there is an asymmetry between invasion of ancestrally XY and ancestrally ZW systems because, e.g., haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system.

@@

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
A full description of our model, including recursion equations, is given in the Appendix. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which assumes a polygamous mating system. 
Gametic competition in monogamous mating systems is, however, equivalent to meiotic drive in our model (described below), as both only alter the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci. %(and the \textbf{M} genotype of their mother in the case of maternal control).
Diploid males and females then experience selection, with relative fitnesses $w_{AA}^{\Hermaphrodite}$, $w_{Aa}^{\Hermaphrodite}$, and $w_{aa}^{\Hermaphrodite}$. %where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Figure \ref{fig:model_outline}A and Table \ref{tab:chisubstitutions}). 
Individuals that are heterozygous at the \textbf{A} locus may experience meiotic drive; a gamete produced by $Aa$ heterozgotes of sex $\Hermaphrodite$ bear allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.

@@

The model outlined above describes both ancestrally-XY and ancestrally-ZW sex-determination systems if we relabel the two sexes as being ancestrally `heterogametic' or ancestrally `homogametic'. 
Without loss of generality, we primarily refer to the ancestrally heterogametic sex as male and the ancestrally homogametic sex as female.
That is, we describe an ancestral XY sex-determination system but our model is equally applicable to an ancestral ZW sex-determination system (relabelling the ancestrally-heterogametic sex as female and the ancestrally-homogametic sex as male).

@@

The evolution of a new sex-determination system requires that a rare mutant allele at the novel sex-determining locus, $m$, increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially in a number of cases of interest. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males or by females, respectively. 
Furthermore, if the $m$ allele is fully epistatically dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves a quadratic characteristic polynomial,
%\textcolor{red}{Mention the possibility that the other roots yield the leading eigenvalue somewhere.}
%\textcolor{blue}{following the logic above, it seems like we say we go from 8 eqns to 4 eqns by considering only one sex, and then from 4 eqns to 2 eqns by ignoring the ancestral SD allele. why then do we have the two quadratic problem? (b/c the $m$ alleles can arise in either sex?)}
$\lambda^2+ b \lambda + c =0$ \textcolor{blue}{(see Appendix for a discussion of other roots - or Sally's proof!)}.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\chi_{mA}\chi_{ma}$ (see table \ref{tab:haplotype_growth}).
%Here, $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\chi_{mA} \chi_{ma}$, where $\lambda_{mi}$ is the (multiplicative) growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\chi_{mi}$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
Here, $b= - (\lambda_{mA} + \lambda_{ma})+(\chi_{mA} + \chi_{ma})$ and $c = (\lambda_{mA}-\chi_{mA}) (\lambda_{ma}-\chi_{ma}) -\chi_{mA} \chi_{ma}$, where $\lambda_{mi}$ is the multiplicative growth rate of mutant haplotypes on background $i\in\{A,a\}$, without accounting for loss due to recombination, and $\chi_{mi}$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other \textbf{A} locus background in heterozygotes (see Table \ref{tab:haplotype_growth}).
The $\lambda_{mi}$ and $\chi_{mi}$, and thus the spread of the mutant $m$ allele, depend on the frequency of alleles at the \textbf{A} and \textbf{X} loci in the ancestral population. 
In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, and among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males. 
We will consider only equilibrium frequencies of alleles, $\hat{p}^\Hermaphrodite_i$, and Y-bearing male gametes, $\hat{q}$, to ensure the eigenvalues of the invasion analysis are valid.

@@

A neo-Y will never invade an ancestral XY system that already has tight linkage with the locus under selection ($r=0$, for details see supplementary \textit{Mathematica} file). 
%When then neo-Y is also tightly linked ($R=0$) a neo-Y will either remain linked to the $A$ allele or to the $a$ allele and so invasion is given directly by the larger of $\lambda_{mA}$ or $\lambda_{ma}$.
A neo-Y haplotype with the same allele as the ancestral Y is neutral ($\lambda_{ma}=1$) and does not change in frequency.
The other neo-Y haplotype will not spread ($\lambda_{mA}<1$) given that the initial equilibrium is stable. 
Therefore, a neo-Y mutation cannot spread ($\lambda \leq 1$) in an ancestral XY system where selected loci are within the non-recombining region around the SDR.
In essence, through tight linkage with the \textbf{A} locus, the ancestral Y becomes strongly specialized on the allele that has the highest fitness across male haploid and diploid phases. 
Given that the ancestral Y is at this equilibrium, it is not possible for a neo-Y to create males that have higher fitness than the ancestral Y.

@@

%Neo-W alleles, on the other hand, can invade an ancestral XY system under some conditions (given in detail in the appendix). 
%%The full characteristic polynomials are given in the appendix (equations \ref{Binvasion} and \ref{Ainvasion}).
%Briefly, neo-W-$A$ and/or neo-W-$a$ haplotypes can spread when rare in the absence of recombination ($\lambda_{ma}>1$ and/or $\lambda_{mA}>1$), depending on the ancestral sex-ratio and allele frequencies. 
%Below we discuss the main forces determining the spread of these neo-W haplotypes and the impact of recombination for the overall success of the neo-W.
%To simplify our discussion we first outline the potential effects of haploid selection and then consider diploid selection in its absence.

>>>>

%involve sex ratio selection or sex-specific selection (e.g., sexually antagonistic selection).

@@

(1) We explicitly consider selection on loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
%(1) We explicitly consider how selection on very tightly sex-linked loci influences the spread of novel sex-determiners. 
Variation at such loci can favour the spread of new sex-determination systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or unlinked) to the locus under selection, which is not predicted by previous theory. 

@@

We find that associations with haploid selected loci can drive transitions between sex determination systems, without requiring sexually-antagonistic selection in diploids. 
Unexpectedly, with haploid selection, transitions between male and female heterogamety can also evolve where linkage with the sex-determining locus is weakened. 
Furthermore, haploid selection in the heterogametic sex can cause sex ratio biases, which may increase or decrease with the spread of new sex chromosomes. 
Thus, we find that transitions between sex-determination systems cannot be simply predicted by selection to equalise the sex ratio. 

@@

%A surprising result given that other determinants of sex allocation typically experience strong Fisherian sex ratio selection to equalize sex ratios. 

@@

Overall, our models reveal that transitions between sex-determination systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
These results predict conditions under which sex-determination systems are likely to be labile and draw novel connections with sex ratio evolution
%These results might be reflected in the lability with which sex-determination systems evolve. %and for the factors determining the phylogenetic pattern of sex-determination systems. 

@@

Predominant theories accounting for the spread of new sex-determination systems by selection involve fitness differences between sexes (e.g., sexually antagonistic selection) or sex ratio selection \citep{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
\citet{vanDoorn:2007eu,vanDoorn:2010hu} and \citet{Muralidhar2018} have shown that new sex-determining loci can be favoured if they arise in close linkage with a locus that experiences sexual antagonism. 
Tighter linkage allows a stronger favourable association to build up between a male-beneficial allele, and a neo-Y chromosome, for example. 
Such associations can favour a new partially-masculinizing or partially-feminizing allele \citep{Muralidhar2018}, a new master sex-determining gene \citep{vanDoorn:2007eu}, and transitions between male and female heterogamety \citep[trans-GSD transitions, ZW to XY or XY to ZW,][]{vanDoorn:2010hu}. 
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determination locus will develop similar, favourable associations and are expected to hinder the spread of a new sex-determination system. 
%Here we extend these studies by explicitly calculating the the equilibrium allele frequencies of loci that are very tightly linked to the ancestral sex-determining region.

The sex ratio is directly determined by the sex-determination system, and it has therefore been suggested that sex ratio selection is a dominant force in the evolution of sex determination (e.g., \citealt{Bull:1983vi}, p 66-67; \citealt{Beukeboom:2014vb}, Chapter 7). 
`Fisherian' sex ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce \citep{Fisher:1930wy,Charnov:1982wg}.
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female \citep{West:2009we}. 
Thus, if the population sex ratio is biased towards one sex, the average per-individual contribution of genetic material to the next generation from the opposite sex is greater. 
Therefore, a mutant that increases investment in the rarer sex will spread via the higher per-individual contributions made by that sex. 
%The default mode of sex ratio selection is `Fisherian' sex ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
%Given that the sex-determination system can directly affect the sex ratio, we might expect Fisherian sex ratio selection to influence the spread of new sex-determination systems. 
In the case of sex-chromosome evolution, \citet{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased \citep{Hamilton:1967ts}. 
They find that new, unlinked sex-determining loci (masculinizing or feminizing mutations, i.e., neo-Y or neo-W loci) can then spread, which restore an even sex ratio.

@@

% from Bai et al. (2016) ``[Segregation distortion] phenomenon is commonly detected via genetic mapping and has been documented in various species, including mouse (Eversley et al., 2010; Casellas et al., 2012), Drosophila (Phadnis and Orr, 2009; Larracuente and Presgraves, 2012; McDermott and Noor, 2012), Tigriopus (Pritchard et al., 2011), rice (Koide et al., 2012; Reflinur et al., 2014; Xu et al., 2014), maize (Tang et al., 2013), and cotton (Yu et al., 2011; Hulse-Kemp et al., 2015).'' Also more in that paper that might be useful. 
%Drosophila http://www.biorxiv.org/content/early/2017/02/06/106336
Segregation distortion provides putative evidence of haploid selection and can sometimes be attributed to meiotic drive and/or gametic competition \citep{Lalanne2004,Fishman2005,Leppala2008,Leppala2013,Didion2015,Didion2016}.
Where it has been characterized, meiotic drive generally occurs either during the production of male or female gametes only \citep{Ubeda:2005gw,Lindholm:2016cw}.
Gametic competition is also typically sex specific, occurring primarily among male gametes, because there are typically many more pollen/sperm than required for fertilization.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte, and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes are known to cause a response to selection \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
Nevertheless, recent studies have demonstrated that sperm competition in animals can alter haploid allele frequencies and increase offspring fitness \citep{Immler:2014im,Alavioon2017}.

There are various ways by which genes experiencing haploid selection could influence transitions between sex-determination systems. 
If we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
In this respect, we might expect that haploid selection would affect transitions between sex-determination systems in a similar manner to sex-specific diploid selection \citep[as explored by][]{vanDoorn:2007eu,vanDoorn:2010hu}. 
That is, new masculinizing mutations (neo-Y chromosomes) could be favoured via associations with alleles that are beneficial in the male haploid stage. 
On the other hand, sex ratios can also become biased by linkage between the sex-determining region and a locus that harbours genetic variation in haploid fitness. 
For example, there are several known cases of sex ratio bias caused by sex-linked meiotic drive alleles \citep[][Chapter 3]{Burt:2006} or selection among X- and Y-bearing pollen \citep{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
It is not immediately clear how the spread of new sex-determination systems would be influenced by the combination of sex ratio biases and associations between haploid selected loci and sex-determining regions.

@@

We find that the spread of novel sex-determiners is influenced by both Fisherian sex ratio selection and by selection on genetically-associated alleles. 
%We find that sex ratio biases caused by haploid selection can exert Fisherian sex ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
Surprisingly, Fisherian sex ratio selection does not dominate; it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing increasingly biased sex ratios. 
In addition to considering haploid selection, another novel development in our model is that we consider loci that are in very tight linkage with the ancestral sex-determining region. 
Because sex-determining loci are often found within a region of suppressed recombination, there can be a significant number of tightly linked loci. 
We find that loci linked with the ancestral sex-determining region can drive transitions in which the heterogametic sex changes, even when the neo-sex-determining locus is less closely linked to loci under selection (either including haploid selection or not). 
%Such transitions are not favoured in models lacking tight linkage and/or haploid selection. 
%Secondly, we allow sex-specific haploid selection to occur on a locus in tight or loose linkage with the ancestral sex-determining region. 
%In addition, when linkage between the ancestral sex-determining locus and a locus under haploid selection is tight, sex ratio biases can be strong such that there is an asymmetry between invasion of ancestrally XY and ancestrally ZW systems because, e.g., haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system.

@@

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
A full description of our model, including recursion equations, is given in the Appendix. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which assumes a polygamous mating system. 
Gametic competition in monogamous mating systems is, however, equivalent to meiotic drive in our model (described below), as either only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci. %(and the \textbf{M} genotype of their mother in the case of maternal control).
Diploid males and females then experience selection, with relative fitnesses $w_{AA}^{\Hermaphrodite}$, $w_{Aa}^{\Hermaphrodite}$, and $w_{aa}^{\Hermaphrodite}$. %where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Figure \ref{fig:model_outline}A and Table \ref{tab:chisubstitutions}). 
Individuals that are heterozygous at the \textbf{A} locus may experience meiotic drive; a gamete produced by $Aa$ heterozgotes of sex $\Hermaphrodite$ bears allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.

@@

The model outlined above describes both ancestrally-XY and ancestrally-ZW sex-determination systems if we relabel the two sexes as being ancestrally `heterogametic' or ancestrally `homogametic'. 
Without loss of generality, we primarily refer to the ancestrally heterogametic sex as male and the ancestrally homogametic sex as female.
That is, we describe an ancestral XY sex-determination system but our model is equally applicable to an ancestral ZW sex-determination system (relabelling the ancestrally-heterogametic sex as female and the ancestrally-homogametic sex as male and switching the labels of males and females throughout).

@@

The evolution of a new sex-determination system requires that a rare mutant allele at the novel sex-determining locus, $m$, increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially in a number of cases of interest. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males or by females, respectively. 
Furthermore, if the $m$ allele is fully epistatically dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves a quadratic characteristic polynomial,
%\textcolor{red}{Mention the possibility that the other roots yield the leading eigenvalue somewhere.}
%\textcolor{blue}{following the logic above, it seems like we say we go from 8 eqns to 4 eqns by considering only one sex, and then from 4 eqns to 2 eqns by ignoring the ancestral SD allele. why then do we have the two quadratic problem? (b/c the $m$ alleles can arise in either sex?)}
$\lambda^2+ b \lambda + c =0$ \textcolor{blue}{(see Appendix for a discussion of other roots - or Sally's proof!)}.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\chi_{mA}\chi_{ma}$ (see table \ref{tab:haplotype_growth}).
%Here, $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\chi_{mA} \chi_{ma}$, where $\lambda_{mi}$ is the (multiplicative) growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\chi_{mi}$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
Here, $b= - (\lambda_{mA} + \lambda_{ma})+(\chi_{mA} + \chi_{ma})$ and $c = (\lambda_{mA}-\chi_{mA}) (\lambda_{ma}-\chi_{ma}) -\chi_{mA} \chi_{ma}$, where $\lambda_{mi}$ is the multiplicative growth rate of mutant haplotypes on background $i\in\{A,a\}$, without accounting for loss due to recombination, and $\chi_{mi}$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other \textbf{A} locus background in heterozygotes (see Table \ref{tab:haplotype_growth}).
The $\lambda_{mi}$ and $\chi_{mi}$, and thus the spread of the mutant $m$ allele, depend on the frequency of alleles at the \textbf{A} and \textbf{X} loci in the ancestral population. 
In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, and among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males. 
We will consider only equilibrium frequencies of alleles, $\hat{p}^\Hermaphrodite_i$, and Y-bearing male gametes, $\hat{q}$, when calculating the eigenvalues.

@@

We find that a neo-Y can never invade an ancestral XY system that already has tight linkage with the locus under selection ($r=0$, for details see supplementary \textit{Mathematica} file). 
%When then neo-Y is also tightly linked ($R=0$) a neo-Y will either remain linked to the $A$ allele or to the $a$ allele and so invasion is given directly by the larger of $\lambda_{mA}$ or $\lambda_{ma}$.
When $R=0$, a neo-Y haplotype with the same allele as the ancestral Y is neutral ($\lambda_{ma}=1$) and does not change in frequency.
The other neo-Y haplotype will not spread ($\lambda_{mA}<1$) given that the initial equilibrium is stable. 
Therefore, a neo-Y mutation cannot spread ($\lambda \leq 1$,  regardless of $R$) in an ancestral XY system where selected loci are within or very near the non-recombining region around the SDR.
In essence, through tight linkage with the \textbf{A} locus, the ancestral Y becomes strongly specialized on the allele that has the highest fitness across male haploid and diploid phases. 
Given that the ancestral Y is at this equilibrium, it is not possible for a neo-Y to create males that have higher fitness than the ancestral Y.

@@

%Neo-W alleles, on the other hand, can invade an ancestral XY system under some conditions (given in detail in the appendix). 
%%The full characteristic polynomials are given in the appendix (equations \ref{Binvasion} and \ref{Ainvasion}).
%Briefly, neo-W-$A$ and/or neo-W-$a$ haplotypes can spread when rare in the absence of recombination ($\lambda_{ma}>1$ and/or $\lambda_{mA}>1$), depending on the ancestral sex ratio and allele frequencies. 
%Below we discuss the main forces determining the spread of these neo-W haplotypes and the impact of recombination for the overall success of the neo-W.
%To simplify our discussion we first outline the potential effects of haploid selection and then consider diploid selection in its absence.
